---
title: "2-16-22-in-class"
author: "Corbin Charpentier"
date: "2/16/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r echo=TRUE}
library(tidyverse)

rm(list=ls())

pp <- function(...) {
    print(paste0(...))
}

fevs <- as_tibble(read.csv("../data/fev.csv"))

length(unique(fevs$age))
```

Question 1. 

1.1. Fit a linear regression model with fev as response and the smoke indicator as predictor. Assess the assumptions of linearity, constant variance and normality using residual plots.
```{R}
plot(lm(fev ~ factor(smoke), data=fevs))


```

1.2. Linearity?

1.3. Constant variance?

1.4. Normality?

1.5. What is your overall conclusion about validity of the results from the linear regression model?

1.6. Add age as a predictor variable to the model and create the residual plots for this model. How do they compare with the residual plots from the first model?
```{R}
plot(lm(fev ~ factor(smoke) + age, data=fevs))
```

Question 2.

Data: ‘field_goals_sample.csv’      

The data are from a random sample of 100 field goal attempts in National Football League games from 2008.

The variables to be used for this exercise are:

distance: distance from ball placement to goal posts (yards)
GOOD: indicator of whether the attempt was successful

2.1. Compare the proportions of field goals that are successful for attempts at distance 40 yards or less versus more than 40 yards using the large-sample Z-test with significance level 0.05. Report the p-value from the test.

2.2. Perform the test using the chi-square test with and without continuity correction.

2.3. Perform the test using Fisher’s exact test.

2.4. How do the results of the different tests compare?
